\chapter{Introduction}

\epigraph{To thy happy children of the future, those of the past send greetings.}{Alma Mater}

At the University of Illinois at Urbana-Champaign, I fundamentally believe that we have a right to make the university better for all future students.
It is a message etched into our Alma Mater and makes up the DNA of our course staff.
As such, we created the coursebook.
The coursebook is a free and open systems programming textbook that anyone can read, contribute to, and modify for now and forever.
We don't think information should be behind a walled garden, and we truly believe that complex concepts can be explained simply and fully, for anyone to understand.
The goal of this book is to teach you the basics and give you some intuition into the complexities of systems programming.

Like any good book, it isn't complete.
We still have plenty of examples, ideas, typos, and chapters to work on.
If you find any issues, please file an \href{https://github.com/illinois-cs241/coursebook/issues}{issue} or email a list of typos to \href{http://cs241.cs.illinois.edu/staff}{CS 241 course staff}, and we'll be happy to work on it.
We are constantly trying to make the book better for students a year and ten years from now.

This work is based on the original coursebook located at \link{https://github.com/angrave/SystemProgramming/wiki}.
All these peoples' hard work is included in the section below.

Thanks again and happy reading!

-- Bhuvy

\section{Authors}

\lstinputlisting[language=console]{AUTHORS}

