\section{Crash course intro to C}

The only way to start learning C is by starting with hello world. As per the original example that Kernighan and Ritchie proposed way back when, the hello world hasn't changed that much.

\begin{lstlisting}[language=C]
#include <stdio.h>
int main(void) {
    printf("Hello World\n");
    return 0;
}
\end{lstlisting}

\begin{enumerate}
	\item The \keyword{\#include} directive takes the file \keyword{stdio.h} (which stands for \textbf{st}an\textbf{d}ard \textbf{i}nput and \textbf{o}utput) located somewhere in your operating system, copies the text, and substitutes it where the \keyword{\#include} was.
	\item The \keyword{int main(void)} is a function declaration. The first word \keyword{int} tells the compiler what the return type of the function is. The part before the parenthesis (\keyword{main}) is the function name. In C, no two functions can have the same name in a single compiled program, shared libraries are a different touchy subject. Then, what comes after is the parameter list. When we give the parameter list for regular functions \keyword{(void)} that means that the compiler should error if the function is called with any arguments. For regular functions having a declaration like \keyword{void func()} means that you are allowed to call the function like \keyword{func(1, 2, 3)} because there is no delimiter. In the case of \keyword{main}, it is a special function. There are many ways of declaring \keyword{main} but the ones that you will be familiar with are \keyword{int main(void)}, \keyword{int main()}, and \keyword{int main(int argc, char *argv[])}.
	\item \keyword{printf("Hello World\n");} is what we call a function call. \keyword{printf} is defined as a part of \keyword{stdio.h}. The function has been compiled and lives somewhere else on our machine. All we need to do is include the header and call the function with the appropriate parameters (a string literal \keyword{"Hello World\n"}). If you don't have the newline, the buffer will not be flushed. It is by convention that buffered IO is not flushed until a newline.
	\item \keyword{return 0;}. \keyword{main} has to return an integer. By convention, \keyword{return 0} means success and anything else means failure.
\end{enumerate}

\begin{verbatim}
$ gcc main.c -o main
$ ./main
Hello World
$
\end{verbatim}

\begin{enumerate}
	\item \keyword{gcc} is short for the GNU-Compiler-Collection which has a host of compilers ready for use. The compiler infers from the extension that you are trying to compile a .c file
	\item \keyword{./main} tells your shell to execute the program in the current directory called main. The program then prints out hello world
\end{enumerate}


